Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-02-21 ms.tex
rotation salvaged

To isolate the physics of interest, our baseline calculations neglect two potentially important effects: heat conduction and radiative cooling.  Electron heat conduction results in an an additional term on the right hand side of the entropy equation of the form:  
\be \dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T), \ee
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conducitivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering exceeds the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field that is oriented perpendicular to the flow could suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).  We show later in $\S\ref{sec:conductivity}$ that neglecting conductivity results in at most only order-unity errors in the key properties of the solutions.

@@

\subsection{Importance of Heat Conduction}
\label{sec:conductivity}

@@

%\input{figures/eta_tab.tex}

@@

  \subsection{Angular Momentum}
  \label{sec:ang}
  Our models are spherically symmetric and do not include the effects of
  the gas' angular momentum. However, galaxies will in general have some
  net rotation, and the gas will generally have some non-zero angular
  momentum. At some radius, $\rcirc$, the specific angular momentum of
  the gas will equal the specific angular momentum of a circular
  orbit. Inside of this radius our assumptions of no rotation and
  spherical symmetry will break down. But if $\rcirc$ is inside of the
  inner sonic point of the flow, the region where angular momentum
  becomes important will be causally disconnected from the rest of the
  flow, and we would be justified in neglecting angular momentum (at
  least on large scales).

  $\rcirc$ depends on the gas injection radius. The largest
  injection radius is $\rs$. Let $\lrs$ be the gas specific angular
  momentum at $\rs$, then

  \begin{align}
    \rcirc=\frac{\lrs}{G \Mbh}
  \end{align}

  \citet{EmsellemCappellari+:2007a} used two-dimensional kinematic
  data to measure the ratio of ordered to random motion in a sample of
  early type galaxies. More precisely, they measure

  \begin{align}
    \lambda_R=\frac{\left<R|V|\right>}{\left<R\sqrt{V^2+\sigma^2}\right>}
  \end{align}

  Where $R$ is the distance is the distance from the galactic center, $V$ is
  the mean stellar velocity, and $\sigma$ is the velocity
  dispersion. The brackets represent a luminosity weighted average.

  We may rewrite the specific angular momentum at $\rs$ in terms of
  $\lambda_R$.

  \begin{align}
    \lrs\simeq\lambdars \sigma|_{r_{\rm s}}^2 \rs^2
  \end{align}

  Then

  \begin{align}
    \rcirc&\simeq\lambdars^2 \frac{\sigma|_{r_{\rm s}}^2}{G \Mbh} \rs^2\\
    &\simeq\lambdars^2 \frac{\rs^2}{\rsoi} \lsim \lambdars^2 \rsoi
  \end{align}

  The inner sonic point, $\rss$ general occurs at $\simeq 1\%$ of
  $\rsoi$. {\bf AG need to explain this somehow. Also, note in our
    simulations the inner sonic point is not properly resolved.}
  So if $\lambdars\lsim 0.1$, $\rcirc<\rss$. In Fig. 2 of
  \citet{EmsellemCappellari+:2007a} $\lambda_R$ is plotted as function
  of radius for a sample of early type galaxies. On scales of $\sim 0.1
  R_e$ $\lambda_R$ is either less than 0.1 or rapidly declining for most
  of their sample %%need to be clearer on rapidly declining also compare
  %% this scale to the sphere of influence scale.
  This suggests that angular momentum should generally be unimportant.

>>>>

To isolate the physics of interest, our baseline calculations neglects three potentially important effects: heat conduction, radiative cooling, and rotation.  Our justification for neglecting rotation will be discussed further in $\S\ref{sec:rotation}$.  Electron heat conduction results in an an additional term on the right hand side of the entropy equation of the form:  
\be \dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T), \ee
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conducitivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering exceeds the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field that is oriented perpendicular to the flow could suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).  We show later in $\S\ref{sec:conductivity}$ that neglecting conductivity results in at most only order-unity errors in the key properties of the solutions.

@@

\subsection{Heat Conduction}
\label{sec:conductivity}

@@

  \subsection{Angular Momentum}
  \label{sec:ang}

Our spherically symmetric model neglects the effects of angular momentum on the gas evolution.  However, all galaxies possess some net rotation, resulting in centrifugal forces becoming important at some radius $r_{\rm circ} = l_{r_{s}}^{2}/(GM_{\bullet})$, where $l_{r_{s}} = r_{s}\langle V_{\phi}\rangle |_{r_s}$ is the specific angular momentum of stars near the stagnation radius, from which most of the accreted mass originates, and $\langle v_{\phi}\rangle$ is their average azimuthal velocity.

Using two-dimensional kinematic data, \citet{EmsellemCappellari+:2007a} measure the ratio of ordered to random motion in a sample of early type galaxies, which they quantify at each galactic radius $R$ by the parameter
  \begin{align}
    \lambda_R=\frac{\langle R|V|\rangle}{\langle R\sqrt{V^2+\sigma^2}\rangle} \underset{R \ll r_{\rm soi}}\sim \frac{V_{\phi}}{\sigma},
  \end{align}
where $\sigma$ is the velocity dispersion and the brackets indicate a luminosity-weighted average.  The circularization radius of the accretion flow can be written in terms of $\lambda_R$ as
\be
\frac{r_{\rm circ}}{r_{s}} \approx \frac{r_{s}}{r_{\rm soi}}\lambda_{R}^{2} \lesssim \lambda_{R}^{2},
\label{eq:rcirc}
\ee
where we have used the definition $r_{\rm soi} \equiv GM_{\bullet}/\sigma^{2}$ and in the second inequality have assumed that $r_{\rm s} \lesssim r_{\rm soi}$ for the thermally-stable solutions of interest.

\citet{EmsellemCappellari+:2007a} (their Fig.~2) find that $\lambda_R < 0.1$ on scales less that 10 per cent of the galaxy radius, in which case equation (\ref{eq:rcirc}) shows that $r_{\rm circ} \lesssim 0.01r_{s}$.  Since this location lies well inside the inner sonic point of the flow, we conclude that angular momentum will only becomes important for thermally stable flows in a region that is causally disconnected from the rest of the flow, justifying its neglect in calculating quantities such as the mass accretion rate.

@@
